R Course Syllabus

### 2023-04-14 Bookdown Syllabus/Bookdown Syllabus.Rmd
adjust instructor picture size
Add lesson 4 and rework lesson order

---
title: "Degrees Draft Syllabus"
author: "Reggie Gilliard"
date: "11/29/2022"
output:
  bookdown::gitbook:
    lib_dir: assets
    split_by: section
    config:
      toolbar:
        position: static
  bookdown::pdf_book:
    keep_tex: true
  bookdown::html_book:
    css: toc.css
documentclass: book
---

@@

![Reggie Gilliard](./Introduction/instructor.jpg) <br/>

@@

## Lesson 3 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}

@@

```{r l3_1_ex1}
# Print data set
mtcars

@@

```{r l3_2_ex1}
1+1 %>%
  sum(2)
```
The pipe operator makes workflows cleaner while still being easy to follow by removing the need for many intermediate saving steps. To illustrate, here is how one could multiply mpg by 2 and divide cyl by 3 in both base R and dplyr.

```{r l3_2_ex2}
# In base R
# First duplicate the data set
mtcars2 <- mtcars
# Then multiply mpg by 2
mtcars2$mpg <- mtcars$mpg * 2
# Then divide cyl by 3
mtcars2$cyl <- mtcars$cyl/3
print(mtcars2)

@@

```{r l3_2_ex3}
mtcars %>%
  filter(cyl == 6)
```

@@

```{r l3_2_ex4}
mtcars %>%
  filter(gear >= 4)
```

@@

```{r l3_2_ex5}
mtcars %>%
  mutate(make = rownames(mtcars)) %>%
  filter(grepl("Mazda", make))
```

@@

```{r l3_2_ex6}
mtcars %>%
  slice(c(1, 3, 5))
```

@@

```{r l3_2_ex7}
mtcars %>%
  slice(seq(1, nrow(mtcars), by = 2))
```

@@

```{r l3_2_ex8}
# First 5 rows
slice_head(mtcars, n = 5)

@@

```{r l3_2_ex9}
# Not arranged
mtcars

@@

```{r l3_2_ex10}
mtcars %>%
  arrange(desc(mpg))
```

@@

```{r l3_2_ex11}
mtcars %>%
  arrange(mpg)

@@

```{r l3_2_ex12}
mtcars %>%
  arrange(mpg, desc(disp))
```

@@

```{r l3_2_ex13}
mtcars %>%
  select(mpg)
```

@@

```{r l3_2_ex14}
mtcars %>%
  select(mpg, gear, carb)

@@

```{r l3_2_ex15}
mtcars %>%
  select(-c("mpg", "gear", "carb"))
```

@@

```{r l3_2_ex16}
mtcars %>%
  rename(MilesPerGallon = mpg)
```

@@

```{r l3_2_ex17}
mtcars %>%
  rename(MilesPerGallon = mpg,
         Cylinders = cyl,
         Carburetors = carb)
```

@@

```{r l3_2_ex18}
mtcars %>%
  rename(`Miles Per Gallon` = mpg)
```

@@

```{r l3_2_ex19}
mtcars %>%
  mutate(mpg_to_cyl = mpg/cyl)
```

@@

```{r l3_2_ex20}
mtcars %>%
  mutate(mpg_to_cyl = mpg/cyl,
         mpg_to_carb = mpg/carb,
         carb_to_cyl = mpg_to_cyl/mpg_to_carb)
```

@@

```{r l3_2_ex21}
mtcars %>%
  mutate(across(c(cyl, mpg, carb), mean, na.rm = T)) %>%
  select(mpg, cyl, carb)
```

@@

```{r l3_2_ex22}
mtcars %>%
  relocate(disp, hp, .before = mpg)
```

@@

```{r l3_2_ex23}
mtcars %>%
  relocate(mpg, cyl, .after = hp)
```

@@

```{r l3_2_ex24}
# Group mtcars by mpg and calculate the mean cylinders and max carburetors within those groups.
mtcars %>%
  group_by(mpg) %>%
  summarize(cyl = mean(cyl),
            carb = max(carb))
```

@@

```{r l3_2_ex25}
# Create mock data
ids <- c(1, 1, 2, 2)
text <- c("Hi", "I'm", "Reggie", "G.")
text_data <- as.data.frame(list(id = ids, text = text))
head(text_data)

@@

## Lesson 4 - Vectors and lists {.tabset .tabset-fade .tabset-pills}

1. What is a vector, what is a list?
    + What is the difference? 
<!-- (for Reggie: https://stackoverflow.com/questions/8594814/what-are-the-differences-between-vector-and-list-data-types-in-r) -->
2. Why are vectors useful?
3. Vectorized operations

```{r lesson4}

>>>>

---
title: "R Course Syllabus"
output:
  bookdown::gitbook:
    lib_dir: assets
    split_by: section
    config:
      toolbar:
        position: static
  bookdown::pdf_book:
    keep_tex: true
  bookdown::html_book:
    css: toc.css
documentclass: book
---

@@

```{r instructor_pic, out.height = "400px", echo = F}
knitr::include_graphics("./Introduction/instructor.jpg")
```

@@

## Lesson 3 - Vectors {.tabset .tabset-fade .tabset-pills}

1. What is a vector?
    + Types of vectors
      + Atomic Vectors
      + Lists
2. Creating vectors
3. Coercion
4. Using vectors
    + Subsetting
    + Vectorized operations

#### **What is a vector?**

In data analysis we often work with more than one value. It can be useful to analyze a single data point, but it is often much more productive to analyze multiple data points sequentially or at the same time. The question arises: how do we store this data?

In R, the most basic way to store data is in a vector. Technically, you have been working with vectors all along when you have been using R to perform calculations and manipulate strings.

```{r l3_1_1}
is.vector(2)
is.vector("RG")
```

These are vectors of length 1.

A vector is a structure that allows you to store data. In R, vectors come in two types: atomic vectors and lists.

##### **Atomic Vectors**

The two vectors I showed above are atomic vectors. You can confirm whether or not a vector is atomic by using the `is.atomic()` function (although this function should be used sparingly as it will also return TRUE in other cases). Experimenting with the vectors I showed above:

```{r l3_1_2}
x <- 2
is.atomic(x)
y <- "RG"
is.atomic(y)
```

An atomic vector is a vector which contains only data of one type. The types are the ones we discussed in lesson 2: logical, numeric, and character (as well as the other, less common, types complex and raw). In the code above `x` is a numeric vector and `y` is a character vector. A logical vector could also be created:

```{r l3_1_3}
z <- TRUE
is.atomic(z)
```

You can check the type of a vector using the `class()` function:

```{r l3_1_4}
class(x)
class(y)
class(z)
```

##### **Lists**

Lists differ from atomic vectors in that they allow the storage of multiple different data types:

```{r l3_1_5}
newList <- list(1, "RG")
is.vector(newList)
newList
```

Data frames, which we will use throughout lesson 4, are built on lists. Since lists are a little more complicated than atomic vectors, we will focus on atomic vectors in the course, but you should be aware that lists exist and you should experiment with them yourself. Lists make some analytic tasks significantly easier or more efficient.

#### **Creating vectors**

There are many ways to create vectors. R has the `vector()` function. You pass it the `mode` or type of vector you would like to create and the length of the vector you would like to create and it creates that vector for you.

```{r l3_2_1}
char_vec <- vector("character", length = 10)
char_vec
```

The vector created by `vector()` is empty, which means you will need to fill it later.

A more common way of creating vectors in R is with the `c()` function (short for combine). You pass this function a series of values you would like combined into a single vector and it returns a vector for you. For example, here is a character vector with my name:

```{r l3_2_2}
my_name <- c("Reggie", "Gilliard")
my_name
```
`c()` took the two strings that I passed it and combined it into one vector which I assigned to `my_name`. Note that unlike the vectors `x`, `y`, and `z` shown above this vector has more than one element:

```{r l3_2_3}
length(x)
length(my_name)
```

#### **Coercion**

I've said that atomic vectors store data of only one type. What happens if you try to combine data of different types into an atomic vector? Let's give it a try:

```{r l3_3_1}
mixed_vec <- c("I am", 28)
mixed_vec
```

What happened here? At first glance, it appears that we were able to do it; but if you looked closely you may have noticed that the number 28 is actually surrounded by quotation marks. This is because of *Coercion*, which is R forcing all elements of the vector to be of a single type. There is a coercion hierarchy: character > double > integer > logical. In other words, if a vector contains 10 elements and just a single one is a character string, the entire vector will become a character string. See, for example:

```{r l3_3_2}
mixed_vec2 <- 1:9 # Store the numbers 1 to 9 in a vector
mixed_vec2
class(mixed_vec2)
mixed_vec2 <- c(mixed_vec2, "10") # Add a single character string
mixed_vec2
class(mixed_vec2)
```

It is important to be aware of how values will be coerced when working with vectors. You can use coercion to your advantage, but it can also work against you and may lead to unintended consequences (such as introducing missing values).

#### **Using vectors**

Until now we have stored a single number, string, or TRUE/FALSE value in each object we created. But, if we were faced with a more complex problem, we could use vectors instead to solve it. Let's consider the task of figuring out who the tallest student is in a class. First, I'll generate some data. It is not important for you to know how to do this.

```{r l3_4_1}
set.seed(89329)
students <- sample(letters, 10, replace = FALSE)
students
heights <- round(runif(10, 60, 72), 2)
heights
```

Now we have 10 students, identified by letters, and 10 heights, one corresponding to each student.

We could visually determine which student is the tallest by looking at the data and seeing that the 3rd height is the largest.

We could then extract the height of that student from the vector using `[`. R uses the brackets to *subset* vectors, lists and data frames. You enter into the brackets the position (*index*) or name of the value you want and R will return it to you. So if we want the height of the 9th student:

```{r l3_4_2}
heights[9]
```

Or if we wanted to get the height of students 4-6 we could do that in two ways:

```{r l3_4_3}
heights[4:6]
heights[c(4, 5, 6)]
```

But we are interested in the height of the 3rd student, since they are the tallest

```{r l3_4_4}
heights[3]
```

And, to find out the student who has this height:

```{r l3_4_5}
students[3]
```
And, finally, to put it all together:

```{r l3_4_6}
paste("Student", students[3], "is the tallest at", heights[3], "inches")
```

But what if, instead, there were 1000 heights and 1000 students? We would not be able to easily discern which student was the tallest by eye. This is what vectors are useful for. R has many functions that make just this kind of analysis simple when using vectors. For example, R's `which.max()` function will return the position of the maximal value in a vector:

```{r l3_4_7}
(max <- which.max(heights))
heights[max]
```

So instead of eyeballing the vector we've used R to quickly find the maximum. Then we can output the same result above:

```{r l3_4_8}
paste("Student", students[max], "is the tallest at", heights[max], "inches")
```

Now suppose we wanted to know the total number of snacks each student ate and there are two types of snacks--salty and sweet.

```{r l3_4_9}
(salty <- sample(1:10, 10, replace = T))
(sweet <- sample(1:10, 10, replace = T))
```

We could get the number of snacks eaten by each student by going throw the indices of the two vectors one-by-one and summing the values like so:

```{r l3_4_10}
salty[1] + sweet[1]
```
And this works fine for a small vector like this, but what about a vector with 1000 values. R offers *vectorized* operations--operations which can be used on entire vectors rather than one element at a time. Addition is one of those vectorized operations. Rather than adding each element individually, we can add the two vectors in one line:

```{r l3_4_11}
salty + sweet
```
This is much simpler and much more efficient.

* [Codecademy Lesson (beginning from the vectors section)](https://www.codecademy.com/courses/learn-r/lessons/introduction-to-r/exercises/why-r)
* [R for Data Science](https://r4ds.had.co.nz/vectors.html)
* [W3 Schools](https://www.w3schools.com/r/r_vectors.asp)

## Lesson 4 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}

@@

```{r l4_1_ex1}
# Print data set
mtcars

@@

```{r l4_2_ex1}
1+1 %>%
  sum(2)
```
The pipe operator makes workflows cleaner while still being easy to follow by removing the need for many intermediate saving steps. To illustrate, here is how one could multiply mpg by 2 and divide cyl by 3 in both base R and dplyr.

```{r l4_2_ex2}
# In base R
# First duplicate the data set
mtcars2 <- mtcars
# Then multiply mpg by 2
mtcars2$mpg <- mtcars$mpg * 2
# Then divide cyl by 3
mtcars2$cyl <- mtcars$cyl/3
print(mtcars2)

@@

```{r l4_2_ex3}
mtcars %>%
  filter(cyl == 6)
```

@@

```{r l4_2_ex4}
mtcars %>%
  filter(gear >= 4)
```

@@

```{r l4_2_ex5}
mtcars %>%
  mutate(make = rownames(mtcars)) %>%
  filter(grepl("Mazda", make))
```

@@

```{r l4_2_ex6}
mtcars %>%
  slice(c(1, 3, 5))
```

@@

```{r l4_2_ex7}
mtcars %>%
  slice(seq(1, nrow(mtcars), by = 2))
```

@@

```{r l4_2_ex8}
# First 5 rows
slice_head(mtcars, n = 5)

@@

```{r l4_2_ex9}
# Not arranged
mtcars

@@

```{r l4_2_ex10}
mtcars %>%
  arrange(desc(mpg))
```

@@

```{r l4_2_ex11}
mtcars %>%
  arrange(mpg)

@@

```{r l4_2_ex12}
mtcars %>%
  arrange(mpg, desc(disp))
```

@@

```{r l4_2_ex13}
mtcars %>%
  select(mpg)
```

@@

```{r l4_2_ex14}
mtcars %>%
  select(mpg, gear, carb)

@@

```{r l4_2_ex15}
mtcars %>%
  select(-c("mpg", "gear", "carb"))
```

@@

```{r l4_2_ex16}
mtcars %>%
  rename(MilesPerGallon = mpg)
```

@@

```{r l4_2_ex17}
mtcars %>%
  rename(MilesPerGallon = mpg,
         Cylinders = cyl,
         Carburetors = carb)
```

@@

```{r l4_2_ex18}
mtcars %>%
  rename(`Miles Per Gallon` = mpg)
```

@@

```{r l4_2_ex19}
mtcars %>%
  mutate(mpg_to_cyl = mpg/cyl)
```

@@

```{r l4_2_ex20}
mtcars %>%
  mutate(mpg_to_cyl = mpg/cyl,
         mpg_to_carb = mpg/carb,
         carb_to_cyl = mpg_to_cyl/mpg_to_carb)
```

@@

```{r l4_2_ex21}
mtcars %>%
  mutate(across(c(cyl, mpg, carb), mean, na.rm = T)) %>%
  select(mpg, cyl, carb)
```

@@

```{r l4_2_ex22}
mtcars %>%
  relocate(disp, hp, .before = mpg)
```

@@

```{r l4_2_ex23}
mtcars %>%
  relocate(mpg, cyl, .after = hp)
```

@@

```{r l4_2_ex24}
# Group mtcars by mpg and calculate the mean cylinders and max carburetors within those groups.
mtcars %>%
  group_by(mpg) %>%
  summarize(cyl = mean(cyl),
            carb = max(carb))
```

@@

```{r l4_2_ex25}
# Create mock data
ids <- c(1, 1, 2, 2)
text <- c("Hi", "I'm", "Reggie", "G.")
text_data <- as.data.frame(list(id = ids, text = text))
head(text_data)

@@

### 2023-04-14 Draft Syllabus.Rmd
adjust instructor picture size
Add lesson 4 and rework lesson order

---
title: "Degrees Draft Syllabus"
author: "Reggie Gilliard"
date: "11/29/2022"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE

@@

* [Codecademy Lesson](https://www.codecademy.com/courses/learn-r/lessons/introduction-to-r/exercises/why-r)
* [You can read more about the data types that we do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)
* [For a deeper dive into creating custom operators](https://adv-r.hadley.nz/functions.html#infix-functions)

## Lesson 3 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}

>>>>

---
title: "R Course Syllabus"
output:
  html_document:
    toc: TRUE
    toc_depth: 2
    toc_float: TRUE

@@

* [Codecademy Lesson (through the Variables section)](https://www.codecademy.com/courses/learn-r/lessons/introduction-to-r/exercises/why-r)
* [You can read more about the data types that we do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)
* [For a deeper dive into creating custom operators](https://adv-r.hadley.nz/functions.html#infix-functions)

## Lesson 3 - Vectors {.tabset .tabset-fade .tabset-pills}

1. What is a vector?
    + Types of vectors
      + Atomic Vectors
      + Lists
2. Creating vectors
3. Coercion
4. Using vectors
    + Subsetting
    + Vectorized operations

#### **What is a vector?**

In data analysis we often work with more than one value. It can be useful to analyze a single data point, but it is often much more productive to analyze multiple data points sequentially or at the same time. The question arises: how do we store this data?

In R, the most basic way to store data is in a vector. Technically, you have been working with vectors all along when you have been using R to perform calculations and manipulate strings.

```{r l3_1_1}
is.vector(2)
is.vector("RG")
```

These are vectors of length 1.

A vector is a structure that allows you to store data. In R, vectors come in two types: atomic vectors and lists.

##### **Atomic Vectors**

The two vectors I showed above are atomic vectors. You can confirm whether or not a vector is atomic by using the `is.atomic()` function (although this function should be used sparingly as it will also return TRUE in other cases). Experimenting with the vectors I showed above:

```{r l3_1_2}
x <- 2
is.atomic(x)
y <- "RG"
is.atomic(y)
```

An atomic vector is a vector which contains only data of one type. The types are the ones we discussed in lesson 2: logical, numeric, and character (as well as the other, less common, types complex and raw). In the code above `x` is a numeric vector and `y` is a character vector. A logical vector could also be created:

```{r l3_1_3}
z <- TRUE
is.atomic(z)
```

You can check the type of a vector using the `class()` function:

```{r l3_1_4}
class(x)
class(y)
class(z)
```

##### **Lists**

Lists differ from atomic vectors in that they allow the storage of multiple different data types:

```{r l3_1_5}
newList <- list(1, "RG")
is.vector(newList)
newList
```

Data frames, which we will use throughout lesson 4, are built on lists. Since lists are a little more complicated than atomic vectors, we will focus on atomic vectors in the course, but you should be aware that lists exist and you should experiment with them yourself. Lists make some analytic tasks significantly easier or more efficient.

#### **Creating vectors**

There are many ways to create vectors. R has the `vector()` function. You pass it the `mode` or type of vector you would like to create and the length of the vector you would like to create and it creates that vector for you.

```{r l3_2_1}
char_vec <- vector("character", length = 10)
char_vec
```

The vector created by `vector()` is empty, which means you will need to fill it later.

A more common way of creating vectors in R is with the `c()` function (short for combine). You pass this function a series of values you would like combined into a single vector and it returns a vector for you. For example, here is a character vector with my name:

```{r l3_2_2}
my_name <- c("Reggie", "Gilliard")
my_name
```
`c()` took the two strings that I passed it and combined it into one vector which I assigned to `my_name`. Note that unlike the vectors `x`, `y`, and `z` shown above this vector has more than one element:

```{r l3_2_3}
length(x)
length(my_name)
```

#### **Coercion**

I've said that atomic vectors store data of only one type. What happens if you try to combine data of different types into an atomic vector? Let's give it a try:

```{r l3_3_1}
mixed_vec <- c("I am", 28)
mixed_vec
```

What happened here? At first glance, it appears that we were able to do it; but if you looked closely you may have noticed that the number 28 is actually surrounded by quotation marks. This is because of *Coercion*, which is R forcing all elements of the vector to be of a single type. There is a coercion hierarchy: character > double > integer > logical. In other words, if a vector contains 10 elements and just a single one is a character string, the entire vector will become a character string. See, for example:

```{r l3_3_2}
mixed_vec2 <- 1:9 # Store the numbers 1 to 9 in a vector
mixed_vec2
class(mixed_vec2)
mixed_vec2 <- c(mixed_vec2, "10") # Add a single character string
mixed_vec2
class(mixed_vec2)
```

It is important to be aware of how values will be coerced when working with vectors. You can use coercion to your advantage, but it can also work against you and may lead to unintended consequences (such as introducing missing values).

#### **Using vectors**

Until now we have stored a single number, string, or TRUE/FALSE value in each object we created. But, if we were faced with a more complex problem, we could use vectors instead to solve it. Let's consider the task of figuring out who the tallest student is in a class. First, I'll generate some data. It is not important for you to know how to do this.

```{r l3_4_1}
set.seed(89329)
students <- sample(letters, 10, replace = FALSE)
students
heights <- round(runif(10, 60, 72), 2)
heights
```

Now we have 10 students, identified by letters, and 10 heights, one corresponding to each student.

We could visually determine which student is the tallest by looking at the data and seeing that the 3rd height is the largest.

We could then extract the height of that student from the vector using `[`. R uses the brackets to *subset* vectors, lists and data frames. You enter into the brackets the position (*index*) or name of the value you want and R will return it to you. So if we want the height of the 9th student:

```{r l3_4_2}
heights[9]
```

Or if we wanted to get the height of students 4-6 we could do that in two ways:

```{r l3_4_3}
heights[4:6]
heights[c(4, 5, 6)]
```

But we are interested in the height of the 3rd student, since they are the tallest

```{r l3_4_4}
heights[3]
```

And, to find out the student who has this height:

```{r l3_4_5}
students[3]
```
And, finally, to put it all together:

```{r l3_4_6}
paste("Student", students[3], "is the tallest at", heights[3], "inches")
```

But what if, instead, there were 1000 heights and 1000 students? We would not be able to easily discern which student was the tallest by eye. This is what vectors are useful for. R has many functions that make just this kind of analysis simple when using vectors. For example, R's `which.max()` function will return the position of the maximal value in a vector:

```{r l3_4_7}
(max <- which.max(heights))
heights[max]
```

So instead of eyeballing the vector we've used R to quickly find the maximum. Then we can output the same result above:

```{r l3_4_8}
paste("Student", students[max], "is the tallest at", heights[max], "inches")
```

Now suppose we wanted to know the total number of snacks each student ate and there are two types of snacks--salty and sweet.

```{r l3_4_9}
(salty <- sample(1:10, 10, replace = T))
(sweet <- sample(1:10, 10, replace = T))
```

We could get the number of snacks eaten by each student by going throw the indices of the two vectors one-by-one and summing the values like so:

```{r l3_4_10}
salty[1] + sweet[1]
```
And this works fine for a small vector like this, but what about a vector with 1000 values. R offers *vectorized* operations--operations which can be used on entire vectors rather than one element at a time. Addition is one of those vectorized operations. Rather than adding each element individually, we can add the two vectors in one line:

```{r l3_4_11}
salty + sweet
```
This is much simpler and much more efficient.

* [Codecademy Lesson (beginning from the vectors section)](https://www.codecademy.com/courses/learn-r/lessons/introduction-to-r/exercises/why-r)
* [R for Data Science](https://r4ds.had.co.nz/vectors.html)
* [W3 Schools](https://www.w3schools.com/r/r_vectors.asp)

## Lesson 4 - Tidy data and dplyr {.tabset .tabset-fade .tabset-pills}
